Flattering Advice: Avoiding Disappointment as a Driver of Gender Discrimination

### 2024-10-08 manuscript/manuscript.Rmd
Before GD

keywords          : "Gender difference, Advice, Interpresonal relationship"
wordcount         : "4,362"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa7"
csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
classoption       : "man"
header-includes:
   - \usepackage{setspace, appendix, placeins, booktabs, tikz, siunitx}
   - \usepackage{tabularx, epigraph, float, colortbl, tabu, pdflscape}
   - \usepackage{tabularray}
   - \usetikzlibrary{positioning}
   - \raggedbottom
   - \usepackage[maxfloats=256]{morefloats}
   - \maxdeadcycles=1000
bibliography      : bibliography.bib
output: 
  # papaja::apa6_docx
  papaja::apa6_pdf
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
always_allow_html: true
---

@@

Seeking and giving advice can present conversational challenges that are
difficult to navigate. Advice seekers may be embarrassed to ask for help
[@Benabou2022], and they may fail to do so if the information they
receive could lead to disappointment [@Gill2012]. Indeed, people prefer
advisers who withhold unpleasant information [@Shalvi2019] and punish
those who tell them things they do not want to hear [@John2019]. People
appear to engage in strategic selection of advice to avoid unpleasant
information and protect the utility they get from positive beliefs about
themselves and the future they anticipate [@Loewenstein2018;
@Loewenstein1987]. But do advice givers anticipate and consider the
hedonic cost of honest information to the recipient? While previous work
has emphasized the instrumental value of advice (e.g., satisfaction with
a decision or accuracy of a prediction), we focus on the motivations of
the person providing advice.

Research on advice has largely focused on settings that do not tackle
belief utility. For example, participants may have been asked to
estimate the number of balls in an urn or choose from pairs of lotteries
[@Benjamin2015]. Advisers have been shown to be motivated to give
accurate advice [@Jonas2003], and engage in perspective-taking to meet
the preferences of the advisee [@Kray2000; @Kray1999]. When they are
uncertain, they even prefer that advisees do not rely (extensively) on
their recommendations [@Ache2020]. Thus, it seems that people are
concerned about providing good advice even when they are not
incentivized for the recipient's decision quality.

In many real settings, however, advice incorporates beliefs about the
recipient. Someone recommending an expensive restaurant conveys that
they think the advisee is wealthy enough to afford an expensive meal,
and someone recommending a student pursue a PhD in economics assumes
that they are intellectually capable. Conversely, telling a poorly
performing student to switch to a less-demanding major conveys a belief
that the student is not capable of improving. Thus, while the painful
truth may prevent someone from making a bad decision, it may come at a
hedonic cost to the recipient and damage the relationship between the
adviser and advisee. We propose that advisers take these costs into
account, even when the interactions are entirely anonymous and they are
incentivized for the quality of the advisee's decision.

@@

In a large, preregistered experiment (n = 1,200), we show that advisers
consider the expectations of advisees, even when such expectations do
not provide informative insights for the underlying decision-making
problem. We find that as a result, showing expectations (but not gender)
to advisers leads men to receive more aspirational advice than women.
Incorporating expectations leads to worse advice, even when advisers are
incentivized based on the outcome of the advisee's decision. In a second
preregistered experiment (n = 700), we show that advisers respond to
incentives for interpersonal considerations. They tell people that they
are more attractive than they believe them to be, and inflate to a
greater extent when incentivized for likability rather than accuracy.
Notably, they inflate even when only incentivized for accuracy. Advisees
indeed reward such flattering advice, rating the advisers more likeable
and no less trustworthy than those who provide more accurate advice.

>>>>

keywords          : "Gender difference, Advice, Interpresonal relationship"
wordcount         : "6,104"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa7"
csl               : "`r system.file('rmd', 'apa7.csl', package = 'papaja')`"
classoption       : "man"
header-includes:
   - \usepackage{setspace, appendix, placeins, booktabs, tikz, siunitx}
   - \usepackage{tabularx, epigraph, float, colortbl, tabu, pdflscape}
   - \usepackage{tabularray}
   - \usetikzlibrary{positioning}
   - \raggedbottom
   - \usepackage[maxfloats=256]{morefloats}
   - \maxdeadcycles=1000
bibliography      : bibliography.bib
output: 
  # papaja::apa6_docx
  papaja::apa6_pdf
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
always_allow_html: true
---

@@

We frequently seek and offer advice, often with the hope of solving problems 
more effectively. Whether we are navigating personal dilemmas or making 
professional decisions, advice is seen as a valuable tool for improving 
outcomes. For instance, in some studies, participants are asked to estimate 
quantities or choose between lotteries [@Benjamin2015]. Advisors are often 
motivated to provide accurate advice [@Jonas2003], and engage in perspective-taking
to align with the preferences of the advisee [@Kray2000; @Kray1999]. When 
uncertain, they may even prefer that advisees rely less on their recommendations
[@Ache2020]. This suggests that people are generally concerned with offering 
good advice, assuming accuracy is the primary objective.

Indeed, seeking and giving advice often involve complex conversational dynamics 
that can be challenging to manage. People frequently struggle with 
decisions about whom to approach for advice, whether they seek it too 
often or too little, and the potential consequences of asking for advice—or
failing to—when it’s expected [@lee1997;@minson2012;@see2011]. 
From the advisor's perspective, factors such as knowledge and expertise [@sniezek2001],
confidence [@sah2013], age [@zhang2020], and communication style [@yaniv2012]
also play critical roles in determining whether accurate advice is 
effectively conveyed.

While much of the advice literature emphasizes the instrumental value of 
advice—such as improving decision satisfaction or prediction accuracy, assuming
both parties prioritize accuracy—research suggests that solving the problem is 
not the only concern in these interactions. Advice seekers may feel embarrassed 
to ask for help [@Benabou2022] and may avoid seeking advice if they fear the 
information could lead to disappointment [@Gill2012]. Moreover, people tend 
to favor advisors who withhold unpleasant information [@Shalvi2019] and even 
punish those who tell them things they would rather not hear [@John2019] This 
indicates that individuals strategically select advice to shield themselves 
from uncomfortable truths, protecting the positive beliefs they hold about 
themselves and their anticipated future [@Loewenstein2018;
@Loewenstein1987]. But do advice givers recognize and account 
for the emotional cost of delivering honest information to the recipient?

In many real settings, advice often incorporates beliefs about the
recipient. Someone recommending an expensive restaurant conveys that
they think the advisee is wealthy enough to afford an expensive meal,
and someone recommending a student pursue a PhD in economics assumes
that they are intellectually capable. Conversely, telling a poorly
performing student to switch to a less-demanding major conveys a belief
that the student is not capable of improving. Thus, while the painful
truth may prevent someone from making a bad decision, it may come at a
hedonic cost to the recipient and damage the relationship between the
adviser and advisee. We propose that advisers take these costs into
account, even when the interactions are entirely anonymous and they are
incentivized for the quality of the advisee's decision.

@@

In a large, preregistered experiment (n = 1,200), we demonstrate that 
advisors take into account advisees' expectations, even when those expectations
offer no relevant insights for solving the decision-making problem. 
Advisors were more likely to encourage advisees to compete against strong
competitors when they saw that the advisees had high expectations of their 
own abilities, compared to when the advisees had low expectations. In a second
preregistered experiment (n = 700), we show that advisors respond to interpersonal
incentives, telling advisees they are more attractive than they believe, and 
inflating their advice more when incentivized for likability rather than accuracy.
Notably, they still inflate their advice even when incentivized solely for accuracy.
Advisees, in turn, reward this flattering advice, rating these advisors as more 
likable and no less trustworthy than those who provide more accurate feedback. 
In the third preregistered experiment (n = 2,000), we replicate the finding that
advisors consider advisees' expectations when giving advice. Additionally, we find
that when only expectations (but not gender) are shown, men receive more aspirational
advice than women. This tendency to factor in expectations ultimately leads to poorer
advice, even when advisors are incentivized based on the outcome of the advisee’s 
decision.
